Equation-of-state insensitive relations after GW170817

### 2019-02-23 paper/bin-love-after-gw170817.tex
katerina comments intro

%Intro
The thermodynamic relation between pressure and density in cold, supranuclear matter, the so-called equation of state (EoS), remains an unsolved problem in nuclear physics. The EoS is critical to our understanding of neutron stars (NSs) because it determines many NS observables, such as their mass, radius, moment of inertia (I), quadrupole moment (Q) and tidal deformability (or Love number). Unfortunately, terrestrial experiments can only probe the EoS to around nuclear saturation density ($\rho_{\text{sat}} \approx 2.5 \times 10^{14} \text{ g/cm}^3$)~\cite{Li:HeavyIon,Tsang:SymmetryEnergy,Centelles:NeutronSkin,Li:CrossSections,Chen:SymEnergy}. Although some temperature-dependent heavy-ion collision experiments are beginning to probe higher densities~\cite{Danielewicz:2002pu}, astrophysical observations of NSs remain ideal for constraining the EoS of cold and ultra dense, nuclear matter.

@@

%reparameterization of the template
The GWs emitted by binary NSs in the late inspiral must then depend on the \emph{tidal deformabilities} $\Lambda_1$ and $\Lambda_2$, which control the linear response of the star's quadrupole deformation to the (electric-type, quadrupole) tidal field of the companion (to leading order in a post-Newtonian expansion~\cite{Blanchet:2013haa})~\cite{Flanagan2008,Vines:2011ud}. These parameters, however, enter the GW Fourier model multiplied by the same function of the GW frequency (at leading post-Newtonian order), making them degenerate, and thus, very difficult to estimate independently with current GW data~\cite{Wade:tidalCorrections}. Instead, one can extract certain combinations of the tidal deformabilities, like a certain mass-weighted tidal deformability $\tilde{\Lambda}$~\cite{Favata:2013rwa,Wade:tidalCorrections}, or one can extract the (mass-independent) coefficients $(\lambda_{0},\lambda_{1},\ldots)$ of a Taylor expansion of the tidal deformabilities about some fiducial mass~\cite{delPozzo:TaylorTidal,Yagi:binLove}. Current detectors are not sensitive enough to accurately measure any of these coefficients, but future detectors will and the information from multiple events could then be combined, since the Taylor expansion coefficient should be common to all events.

%Universal relations
Lacking enough sensitivity in current GW observations, one is forced to only estimate the mass-weighted tidal deformability, but this prevents the independent extraction of $\Lambda_{1}$ and $\Lambda_{2}$. Yagi and Yunes~\cite{Yagi:binLove} solved this problem by finding ``approximately universal'' or ``EoS-insensitive'' relations between the symmetric and anti-symmetric combinations of the tidal deformabilities $\Lambda_{s,a}=\frac{1}{2}(\Lambda_1 \pm \Lambda_2)$, the so-called ``binary Love relations." These relations can be used to analytically express $\Lambda_{s}$ in terms of $\Lambda_{a}$ (or vice-versa), making the mass-weighted tidal deformability a function of only $\Lambda_{a}$. A measurement of the mass-weighted tidal deformability then implies a measurement of $\Lambda_{a}$, and through the use of the binary Love relations, also a measurement of $\Lambda_{s}$, which then allows for the inference of the individual tidal deformabilities $\Lambda_{1}$ and $\Lambda_{2}$~\cite{Yagi:binLove}. With those at hand, one can further use EoS-insensitive relations between the tidal deformabilities and the compactness, the so-called ``C-Love relations''~\cite{Yagi:ILQ}, to infer the radii of the NSs, and thus, to place two constraints in the mass-radius plane, one for each star in the binary~\cite{Yagi:binLove}. This idea was recently implemented in GW170817, allowing for the first EoS-independent constraints on the mass-radius curve using GW data~\cite{LIGO:posterior}.

EoS-insensitive relations can in fact be used for more than just measuring the nuclear EoS. For years, the theoretical physics community considered the possibility of using measurements of NS properties, like the mass, the radius and the moment of inertia, to constrain deviations from General Relativity in the strong-field regime. Certain modified theories of gravity, like scalar tensor theories with spontaneous scalarization~\cite{Damour:1996ke}, Einstein-Aether and Horava gravity~\cite{Eling:2007xh,Yagi:2013ava,Yagi:2013qpa}, dynamical Chern-Simons gravity~\cite{Yunes:2009ch}, beyond Horndesky theories~\cite{Sakstein:2016oel}, modify such NS observables, but unfortunately, these modifications are typically degenerate with the nuclear EoS. Yagi and Yunes solved this problem by finding EoS-insensitive relations between the moment of inertia, the tidal deformability (or Love number) and the quadrupole moment, the so-called ``I-Love-Q'' relations~\cite{Yagi:ILQ}. Given a measurement of the Love number for a given NS, for example through GW observations, the I-Love-Q relations can be used to infer the moment of inertia or the quadrupole moment. A second independent measurement of either of these two quantities, for example through binary pulsar observations~\cite{Lattimer:2004nj} or observations with the Neutron star Interior Composition ExploreR (NICER)~\cite{Ozel:2015ykl}, then allows an EoS-insensitive test of General Relativity in the strong field regime~\cite{Yagi:ILQ,Gupta:2017vsl}.

@@

%How we improved this
We begin by investigating the increase in EoS insensitivity due to the constraints placed by GW170817 on the allowed space of EoSs. We take the 90\% credible region constructed from the posterior probability distributions on the pressure-density plane~\cite{LIGO:posterior} obtained from GW170817~\cite{TheLIGOScientific:2017qsa} as a prior on the set of EoSs that is compatible with this observation. We then generate two large samples of spectral EoSs~\cite{Lindblom:2018rfr}, one in which this prior is enforced (the ``constrained'' sample) and another in which the prior is not enforced (the ``unconstrained'' sample). We repeat the analysis done by Yagi and Yunes~\cite{Yagi:binLove,Yagi:ILQ} on both sets of EoSs and find that the EoS-insensitive relations present less EoS variability with the constrained set. In particular, the EoS-insensitivity increases by a factor of $\sim 60$\% in the binary Love relations (for stars with mass ratio larger than $0.75$), by a factor of $\sim 70\%$ in the C-Love relations, and by factors of $\sim 50$\% in the I-Love-Q relations (see Table~\ref{tab:maxVar} and Fig.~\ref{fig:ILQ} for more details).

@@

Last for not least, we study the importance of using the improved EoS-insensitive relations in future GW observations. The use of EoS-insensitive relations introduces systematic uncertainties in parameter estimation because of the intrinsic non-zero EoS-variability in these relations, which one must marginalize over. These uncertainties are currently irrelevant because statistical uncertainties in parameter estimation are much larger with current detectors. But as the detector sensitivity is improved, the signal-to-noise ratio and the number of events that will be detected will increase, therefore decreasing the statistical uncertainties below the systematic one due to EoS variability. We carry out a simple Fisher analyses to estimate when the statistical uncertainties become comparable to the systematic uncertainties due to EoS variability and find that this occurs for Voyager-class detectors.

Figure~\ref{fig:stackedFisher} shows this result in more detail. We here present the (Fisher-estimated) statistical uncertainty in the measurement of $\lambda_{0}$ with various detectors (LIGO O2~\cite{aLIGO}, Advanced LIGO at design sensitivity (aLIGO)~\cite{aLIGO}, LIGO A\texttt{+} (A\texttt{+})\cite{Ap_Voyager_CE}, Voyager~\cite{Ap_Voyager_CE}, the Einstein Telescope (ET)~\cite{ET}, and the Cosmic Explorer (CE)~\cite{Ap_Voyager_CE}) for an event identical to GW170817. The x-axis shows the signal-to-noise ratio for a GW170817 event detected with each of these detectors. We also present the combined statistical uncertainty $\sigma^A_N$ after $N$ detections with each of these instruments, with the top and bottom of the region representing the most optimistic and pessimistic expectation for the number of detections expected from the binary NS merger rate (see also Table~\ref{tab:variances}). These statistical uncertainties should be compared to the systematic uncertainty in $\lambda_{0}$ due to EoS-variability improved with the constrained set. Observe that the statistical and the systematic uncertainties cross for Voyager-class detectors.

>>>>

%Intro
The thermodynamic relation between pressure and density in cold, supranuclear matter, the so-called equation of state (EoS), remains an unsolved problem in nuclear physics. The EoS is critical to our understanding of neutron stars (NSs) because it determines many NS observables, such as their mass, radius, moment of inertia (I), quadrupole moment (Q) and tidal deformability (or Love number). Unfortunately, terrestrial experiments can only probe the EoS to around nuclear saturation density ($\rho_{\text{sat}} \approx 2.5 \times 10^{14} \text{ g/cm}^3$)~\cite{Li:HeavyIon,Tsang:SymmetryEnergy,Centelles:NeutronSkin,Li:CrossSections,Chen:SymEnergy}. Although some temperature-dependent heavy-ion collision experiments can probe higher densities~\cite{Danielewicz:2002pu}, astrophysical observations of NSs remain ideal for constraining the EoS of cold and ultra dense, nuclear matter.

@@

%reparameterization of the template
The GWs emitted by binary NSs in the late inspiral must then depend on the \emph{tidal deformabilities} $\Lambda_1$ and $\Lambda_2$, which control the linear response of the star's quadrupole deformation to the (electric-type, quadrupole) tidal field of the companion (to leading order in a post-Newtonian expansion~\cite{Blanchet:2013haa})~\cite{Flanagan2008,Vines:2011ud}. \sout{These parameters, however, enter the GW Fourier model multiplied by the same function of the GW frequency} \kc{that's not correct, the "function" depends on the mass ratio} These parameters, however, enter the GW Fourier model at the same post-Newtonian order, making them degenerate, and thus, very difficult to estimate independently with current GW data~\cite{Wade:tidalCorrections}. Instead, one can extract certain combinations of the tidal deformabilities, such as a certain mass-weighted tidal deformability $\tilde{\Lambda}$~\cite{Favata:2013rwa,Wade:tidalCorrections}, or one can extract the (mass-independent) coefficients $(\lambda_{0},\lambda_{1},\ldots)$ of a Taylor expansion of the tidal deformabilities about some fiducial mass~\cite{delPozzo:TaylorTidal,Yagi:binLove}. Current detectors are not sensitive enough to accurately measure any of these coefficients, but future detectors will and the information from multiple events could then be combined, since the Taylor expansion coefficient should be common to all events.

%Universal relations
Lacking enough sensitivity in current GW observations, one is forced to only estimate the mass-weighted tidal deformability, but this prevents the independent extraction of $\Lambda_{1}$ and $\Lambda_{2}$. Yagi and Yunes~\cite{Yagi:binLove} \sout{solved} \kc{proposed a solution to this problem} this problem by finding ``approximately universal'' or ``EoS-insensitive'' relations between the symmetric and anti-symmetric combinations of the tidal deformabilities $\Lambda_{s,a}=\frac{1}{2}(\Lambda_1 \pm \Lambda_2)$, the so-called ``binary Love relations." These relations can be used to analytically express $\Lambda_{s}$ in terms of $\Lambda_{a}$ (or vice-versa), making the mass-weighted tidal deformability a function of only $\Lambda_{a}$. A measurement of the mass-weighted tidal deformability then implies a measurement of $\Lambda_{a}$, and through the use of the binary Love relations, also a measurement of $\Lambda_{s}$, which then allows for the inference of the individual tidal deformabilities $\Lambda_{1}$ and $\Lambda_{2}$~\cite{Yagi:binLove}. With those at hand, one can further use EoS-insensitive relations between the tidal deformabilities and the compactness, the so-called ``C-Love relations''~\cite{Yagi:ILQ}, to infer the radii of the NSs, and thus, to place two constraints in the mass-radius plane, one for each star in the binary~\cite{Yagi:binLove}. This idea was recently implemented in GW170817, allowing for an EoS-independent constraints on the mass-radius curve using GW data~\cite{Katerina:residuals,LIGO:posterior}.

EoS-insensitive relations can in fact be used for more than just measuring the nuclear EoS. For years, the theoretical physics community considered the possibility of using measurements of NS properties, such as the mass, the radius and the moment of inertia, to constrain deviations from General Relativity in the strong-field regime. Certain modified theories of gravity, such as scalar tensor theories with spontaneous scalarization~\cite{Damour:1996ke}, Einstein-Aether and Horava gravity~\cite{Eling:2007xh,Yagi:2013ava,Yagi:2013qpa}, dynamical Chern-Simons gravity~\cite{Yunes:2009ch}, beyond Horndesky theories~\cite{Sakstein:2016oel}, modify such NS observables, but unfortunately, these modifications are typically degenerate with the nuclear EoS. Yagi and Yunes \sout{solved} \kc{proposed to solve} this problem by finding EoS-insensitive relations between the moment of inertia, the tidal deformability (or Love number) and the quadrupole moment, the so-called ``I-Love-Q'' relations~\cite{Yagi:ILQ}. Given a measurement of the Love number for a given NS, for example through GW observations, the I-Love-Q relations can be used to infer the moment of inertia or the quadrupole moment. A second independent measurement of either of these two quantities, for example through binary pulsar observations~\cite{Lattimer:2004nj} or observations with the Neutron star Interior Composition ExploreR (NICER)~\cite{Ozel:2015ykl}, then allows an EoS-insensitive test of General Relativity in the strong field regime~\cite{Yagi:ILQ,Gupta:2017vsl}.

@@

%How we improved this
We begin by investigating the increase in EoS insensitivity due to the constraints placed by GW170817 on the allowed space of EoSs. We take the 90\% credible region constructed from the posterior probability distributions on the pressure-density plane~\cite{Carney:2018sdv,LIGO:posterior} obtained from GW170817~\cite{TheLIGOScientific:2017qsa} as a prior on the set of EoSs that is compatible with this observation. We then generate two large samples of spectral EoSs~\cite{Lindblom:2018rfr}, one in which this prior is enforced (the ``constrained'' sample) and another in which the prior is not enforced (the ``unconstrained'' sample). We repeat the analysis done by Yagi and Yunes~\cite{Yagi:binLove,Yagi:ILQ} on both sets of EoSs and find that the EoS-insensitive relations present less EoS variability with the constrained set. In particular, the EoS-insensitivity increases by a factor of $\sim 60$\% in the binary Love relations (for stars with mass ratio larger than $0.75$), by a factor of $\sim 70\%$ in the C-Love relations, and by factors of $\sim 50$\% in the I-Love-Q relations (see Table~\ref{tab:maxVar} and Fig.~\ref{fig:ILQ} for more details).

@@

Last but not least, we study the importance of using the improved EoS-insensitive relations in future GW observations. The use of EoS-insensitive relations introduces systematic uncertainties in parameter estimation because of the intrinsic non-zero EoS-variability in these relations, which one must marginalize over. These uncertainties are currently irrelevant because statistical uncertainties in parameter estimation are much larger with current detectors. But as the detector sensitivity is improved, the signal-to-noise ratio and the number of events that will be detected will increase, therefore decreasing the statistical uncertainties below the systematic one due to EoS variability. We carry out a \sout{simple} Fisher analyses to estimate when the statistical uncertainties become comparable to the systematic uncertainties due to EoS variability and find that this occurs for Voyager-class detectors.

Figure~\ref{fig:stackedFisher} shows this result in more detail. We present the (Fisher-estimated) statistical uncertainty in the measurement of $\lambda_{0}$ with various detectors (LIGO O2~\cite{aLIGO}, Advanced LIGO at design sensitivity (aLIGO)~\cite{aLIGO}, LIGO A\texttt{+} (A\texttt{+})\cite{Ap_Voyager_CE}, Voyager~\cite{Ap_Voyager_CE}, the Einstein Telescope (ET)~\cite{ET}, and the Cosmic Explorer (CE)~\cite{Ap_Voyager_CE}) for an event \sout{identical to} similar to GW170817. The x-axis shows the signal-to-noise ratio for a GW170817 event detected with each of these detectors. We also present the combined statistical uncertainty $\sigma^A_N$ after $N$ detections with each of these instruments, with the top and bottom of the region representing the most optimistic and pessimistic expectation for the number of detections expected from the binary NS merger rate \kc{add citation for the merger rate} (see also Table~\ref{tab:variances}). These statistical uncertainties should be compared to the systematic uncertainty in $\lambda_{0}$ due to EoS-variability improved with the constrained set. We find that the statistical and the systematic uncertainties cross for Voyager-class detectors.
